---
title: "Vaccination status among measles cases"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's make the following assumptions:

* the age distribution for the children of age considered in this study is flat;
* the exposure to measles is independent of age;
* the susceptibility is determined by vaccination only;
* the prevalence of chronic disease is independent of age;
* the vaccine efficacies are independent of the fact that children have chronic
disease of not;

A child can be infected either because s/he's not vaccinated or because s/he's
vaccinated and the vaccine does not work:

$$
M = [(1 - p) + (1 - e) p] N
$$

where $M$ is the total number of measles cases, $p$ is the vaccine coverage,
$e$ is the vaccine efficacy and $N$ is the total number of children in the
population. From this equation, we can derive the proportion of children that
are vaccined amond the measles cases:

$$
P = \frac{(1 - e) p}{(1 - p) + (1 - e) p}
$$

Let's now consider the proportions of children that are vaccinated in the 4
categories of children as below:

```{r, eval = FALSE}
             | 9 < age < 18 | 18 < age
-------------|--------------|---------
non-chronic  |      P1      |    P2
chronic dis. |      P3      |    P4
```

Then we have 4 equations with 4 unknown:

$$
P_1 = \frac{(1 - e_1) p_1}{(1 - p_1) + (1 - e_1) p_1}
$$
$$
P_2 = \frac{(1 - e_2) p_1}{(1 - p_1) + (1 - e_2) p_1}
$$
$$
P_3 = \frac{(1 - e_1) p_2}{(1 - p_2) + (1 - e_1) p_2}
$$
$$
P_4 = \frac{(1 - e_2) p_2}{(1 - p_2) + (1 - e_2) p_2}
$$

Where $p_1$ and $p_2$ are the vaccine coverages of children without and with
chronic disease respectively and $e_1$ and $e_2$ are the vaccine efficiencies
after the first and second doses respectively. We can estimate the parameters
$p_1$, $p_2$, $e_1$, $e_2$ by maximum likelihood, using a binomial distribution
of errors. Let's look at the implementation in R with the package `bbmle`:

```{r}
library(bbmle)
```

First we need 2 functions to transform the parameter values in order to order
to constraint the parameters between 0 and 1. The first function is the logit
function:

```{r}
logit <- function(x) {
  log(x / (1 - x))
}
```

And the second function is the inverse logit function:

```{r}
inverse_logit <- function(x) {
  exp(x)/ (1 + exp(x))
}
```

The we need a function that computes the minus log-likelihood as a function of
the parameters values and for some given values of the data V1, V2, V3, V4 (the
numerators of P1, P2, P3 and P4 respectively), and N1, N2, N3 and N4 (the
denominators of P1, P2, P3 and P4 respectively):

```{r}
mLL <- function(p1, p2, e1, e2) {
  p1 <- inverse_logit(p1)
  p2 <- inverse_logit(p2)
  e1 <- inverse_logit(e1)
  e2 <- inverse_logit(e2)
  - sum(dbinom(c(V1, V2, V3, V4), c(N1, N2, N3, N4),
               c(((1 - e1) * p1) / ((1 - p1) + (1 - e1) * p1),
                 ((1 - e2) * p1) / ((1 - p1) + (1 - e2) * p1),
                 ((1 - e1) * p2) / ((1 - p2) + (1 - e1) * p2),
                 ((1 - e2) * p2) / ((1 - p2) + (1 - e2) * p2)), TRUE))
}
```

From here, let's define the function that puts initial values of the parameters
and data together and does the maximum likelihood estimation:

```{r}
mlestimation <- function(p1, p2, e1, e2, N1, N2, N3, N4, V1, V2, V3, V4, ...) {
  mle2(mLL, as.list(logit(c(p1 = p1, p2 = p2, e1 = e1, e2 = e2))),
       data = list(N1 = N1, N2 = N2, N3 = N3, N4 = N4, V1 = V1, V2 = V2, V3 = V3, V4 = V4), ...)
}
```

Let's try it:

```{r}
out <- mlestimation(.5, .5, .9, .9, 526, 356, 119, 126, 57, 121, 8, 12)
```

Which gives the following estimates:

```{r}
inverse_logit(coef(out))
```

Suggesting that the vaccine coverage for children without chronic diseases would
be around 73% and that the vaccine coverage of children with chronic diseases
would be around 44% in comparison. It also strangely suggest that the vaccine
efficiency after the first dose (95%) would be higher than after the second dose
(82%). Let's see the confidence intervals:

```{r}
confint(out)
```

