---
title: "Model"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

## A simple 2-doses vaccination model

### Demographics:

Let's consider an age cohort from 0 to $t_3$ day-old and let's assume that
contacts within this cohort are homogeneous. Let's assume that the exact same
number of babies are born every day. In consequence, the number of children of
a given age (in days) represents $n = 1 / d$ percents of the total number of
children in the 0-$t_3$ day-old cohort.

### Vaccination:

Let's consider a vaccination with two doses at $t_1$ and $t_2$ and let's call
$c_1$ the vaccine coverage of the first dose and $c_2$ the vaccine coverage of
the second dose (i.e. given the first dose). Let's call $e_1$ and $e_2$ the
vaccine efficiencies of the first and second doses respectively.

The percentage of the cohort with one dose thus reads:

$$
p_1 = [(t_2 - t_1) \times c_1 + (t_3 - t_2) \times c_1 \times (1 - c_2)] \times n
$$

And the percentage of the cohort with two doses reads:

$$
p_2 = (t_3 - t_2) \times c_1 \times c_2 \times n
$$

The cohort level of sero-protection then reads:

$$
s = p_1 \times e_1 + p_2 \times e_2
$$

### An alternative expression of the vaccine coverage of the second dose

Sometimes the vaccine coverage of the second dose is expressed as $c_2'$ with
the total number of children expected to received the 2 doses of vaccine in the
denominator instead of number of children having already received a first dose
as done for $c_2$, meaning:

$$
c_2' = c_1 \times c_2
$$

Then the above equations have to be updated with the percentage of the cohort
with one dose now reading:

$$
p_1 = [(t_2 - t_1) \times c_1 + (t_3 - t_2) \times (c_1 - c_2')] \times n
$$

Note that we necessarily have the constraint $c_2' \leq c_1$ here. The
percentage of the cohort with two doses now reads:

$$
p_2 = (t_3 - t_2) \times c_2' \times n
$$

The cohort level of sero-protection then reads:

$$
s = p_1 \times e_1 + p_2 \times e_2
$$

### Cohort level of protection

This functions uses the above equations to compute the population level of
sero-protection:

```{r}
protection1 <- function(t1, t2, t3, c1, c2, e1, e2, conditional = TRUE) {
  if (! conditional) {
    if (c2 > c1) stop("c2 cannot be greater than c1.")
    c2 <- c2 / c1
  }
  n <- 1 / t3
  one_dose <- ((t2 - t1) * c1 + (t3 - t2) * c1 * (1 - c2)) * n
  two_doses <- (t3 - t2) * c1 * c2 * n
  one_dose * e1 + two_doses * e2
}
```

Let's see an example of this function use with the two expressions of the
coverage of the second dose:

```{r}
protection1(t1 = 9 * 30,  # timing of the first dose (in days)
            t2 = 18 * 30, # timing of the second dose (in days)
            t3 = 5 * 365, # duration of the cohort (in days)
            c1 = .85,     # vaccine coverage first dose (btwn 93% and 96%)
            c2 = .85,     # vaccine coverage second dose | first dose
            e1 = .85,     # vaccine efficiency first dose
            e2 = .95      # vaccine efficiency second dose
)
```

and:

```{r}
protection1(t1 = 9 * 30,  # timing of the first dose (in days)
            t2 = 18 * 30, # timing of the second dose (in days)
            t3 = 5 * 365, # duration of the cohort (in days)
            c1 = .85,     # vaccine coverage first dose (btwn 93% and 96%)
            c2 = .85,     # vaccine coverage second dose | first dose
            e1 = .85,     # vaccine efficiency first dose
            e2 = .95,     # vaccine efficiency second dose
            conditional = FALSE
)
```

